---
title: "Biofeedback Proof of Concept Study"
output: github_document
date: "2023-09-19"
---

``` {r results='hide', message=FALSE, warning=FALSE, echo=FALSE}
library(tidyr)
library(stats)
library(tidyverse)
library(ggpubr)
library(cowplot)
library(corrplot)
library(psych)
library(dplyr)
library(DescTools)
library(ggrain)
library(knitr)
library(rstatix)

```
# Background

This study provides a proof of concept of two virtual reality mini-games, to both teach participants a physiological regulation technique and subsequently to apply this technique in a stressful biofefeedback environment. 

### Study Design

Participants physiological measurements were taken at baseline (seated at rest) as well as during the training (paced breathing training) and stressor (biofeedback horror). 

The physiological measurements reported here are Heart Rate (HR), heart rate variability as indexed by SDNN, and respiration rate (resp).

Initial analysis was conducted using a paired t-test for the HR difference from baseline to stressor, and Signs tests for the SDNN and respiratory difference from baseline to stressor. 


``` {r echo = FALSE}

setwd("/Users/luciedaniel-watanabe/Desktop/NT:PhD/biofeedback/paper writeup/data")
resp <- read.csv("resp_percentages.csv")
quest <- read.csv("biofeedbackQuestionnaires.csv")

SDNN <- read.csv("SDNN.csv")
HR <- read.csv("HR.csv")
MeanRR <- read.csv("MeanRR.csv")


###cardiac measures are listed with ids from lowest to highest 
##so it's important to make sure that the other dataframes are too
##this is definitely not something I messed up before 
resp <- resp[order(resp$p_id),]
quest <- quest[order(quest$p_id),]


###normalising the SDNN data according to the calculation supplied in 
## de Geus et al 2019, psychophysiology 

SDNNnorm <- data.frame(p_id = SDNN$p_id)
SDNNnorm$SDNN_baseline <- 100 * (SDNN$SDNN_baseline / MeanRR$baseline)
SDNNnorm$SDNN_training <- 100 * (SDNN$SDNN_training / MeanRR$boat)
SDNNnorm$SDNN_stress <- 100 * (SDNN$SDNN_stress / MeanRR$dungeon)

SDNN <- SDNNnorm

###get rid of things we won't use again 
rm(SDNNnorm, MeanRR)

HR <- HR[complete.cases(HR),]
SDNN <- SDNN[complete.cases(SDNN),]


##### Calculating the difference measures

#Calculate the difference in stress rating between stressor and boat
quest$stress_diff <- quest$dungeon_score - quest$boat_score

#Difference in SDNN between training and baseline, and stressor and baseline
SDNN$SDNN_boat_BL_diff <- SDNN$SDNN_training - SDNN$SDNN_baseline
SDNN$SDNN_stress_BL_diff <- SDNN$SDNN_stress - SDNN$SDNN_baseline
SDNN$SDNN_training_stress_diff <- SDNN$SDNN_stress - SDNN$SDNN_training


#Difference in HR between training and baseline, and stressor and baseline
HR$HR_boat_BL_diff <- HR$HR_training - HR$HR_baseline
HR$HR_stress_BL_diff <- HR$HR_stress - HR$HR_baseline
HR$HR_training_stress_diff <- HR$HR_stress - HR$HR_training


options(scipen=999)


##Paired t tests to see whether the difference from baseline to stress conditions is significant 
#t.test(SDNN$SDNN_baseline, SDNN$SDNN_stress, paired = T)
##Paired t tests to see whether the difference from baseline to stress conditions is significant 
t.test(SDNN$SDNN_baseline, SDNN$SDNN_stress, paired = T)
t.test(HR$HR_baseline, HR$HR_stress, paired = T)
t.test(resp$resp_baseline, resp$resp_stress, paired = T)


```

# Plots for Physiological change

## HR and cvSDNN change: baseline, training, stressor

```{r echo = FALSE}

######################################################################
################ PLOTS FOR RESP, HR, AND HRV CHANGE ################ 
##################################################################### 

####This section is all ggplots for the graphs showing respiration, HR and HRV
### change.

cbbPalette <- c(  "#009E73", "#0072B2", "#D55E00")
cbPalette <- c( "#0072B2", "#D55E00")

##ggplot for the heart rate data across the baseline, training and stressor 
gg_HR <- data.frame(value  = c(HR$HR_baseline, HR$HR_training, HR$HR_stress),
                   category = c(rep('baseline', 44), rep('training', 44), rep('stress', 44)),
                   paired = c(1:44, 1:44, 1:44))

gg_HR$category <- factor(gg_HR$category, levels = c("baseline", "training", "stress"))


ggplot(gg_HR, aes(x= category, y=value, fill=category)) +
  geom_boxplot(show.legend = F)+
  # geom_point() is used to make points at data values
  geom_point(show.legend = F)+
  # geom_line() joins the pa"ired datapoints
  scale_fill_manual(values=cbbPalette)+
  ylab("HR") +
  xlab("")+
  ggtitle("HR in Baseline, Training and Stressor")+
  theme(panel.background = element_blank())


###plot for difference in heart rate between the baseline and stressor

gg_HR_stress_bl <- data.frame(value  = c(HR$HR_baseline, HR$HR_stress),
                   category = c(rep('baseline', 44), rep('stress', 44)),
                   paired = c( 1:44, 1:44))



##ggplot for SDNN across the baseline, training and stressor 
gg_SDNN_supp <- data.frame(value  = c(SDNN$SDNN_baseline, SDNN$SDNN_training, SDNN$SDNN_stress),
                           category = c(rep('baseline', 44), rep('training', 44), rep('stressor', 44)),
                           paired = c(1:44, 1:44, 1:44))
gg_SDNN_supp$category <- factor(gg_SDNN_supp$category, levels = c("baseline", "training", "stressor"))

ggplot(gg_SDNN_supp, aes(x= category, y=value, fill=category)) +
  geom_boxplot(show.legend = F)+
  # geom_point() is used to make points at data values
  geom_point(show.legend = F)+
  # geom_line() joins the pa"ired datapoints
  scale_fill_manual(values=cbbPalette)+
  ylab("cvSDNN") +
  xlab("")+
  ggtitle("cvSDNN in Baseline, Training and Stressor")+
  theme(panel.background = element_blank())


```

## HR And cvSDNN for baseline/stressor

``` {r echo = FALSE}
###plot for difference in heart rate between the baseline and stressor

gg_HR_stress_bl <- data.frame(value  = c(HR$HR_baseline, HR$HR_stress),
                              category = c(rep('baseline', 44), rep('stress', 44)),
                              paired = c( 1:44, 1:44))


###save the HR and SDNN plots so they can be combined into one image
HR_plot <- ggplot(gg_HR_stress_bl, aes(x=category, y=value, fill=category)) +
  geom_boxplot(show.legend = F, coef = 0)+
  # geom_point() is used to make points at data values
  geom_point(show.legend = F)+
  scale_fill_manual(values=cbPalette)+
  # geom_line() joins the paired datapoints
  geom_line(aes(group=paired),  alpha = 0.4) +
  ylab("HR") +
  xlab("")+
  ggtitle("HR for Baseline and Stressor")+
  theme(panel.background = element_blank())


gg_SDNN_stress_bl <- data.frame(value  = c(SDNN$SDNN_baseline, SDNN$SDNN_stress),
                                category = c(rep('baseline', 44), rep('stress', 44)),
                                paired = c( 1:44, 1:44))


SDNN_plot <- ggplot(gg_SDNN_stress_bl, aes(x=category, y=value, fill=category)) +
  geom_boxplot(show.legend = F, coef = 0)+
  # geom_point() is used to make points at data values
  geom_point(show.legend = F)+
  scale_fill_manual(values=cbPalette)+
  # geom_line() joins the paired datapoints
  geom_line(aes(group=paired),  alpha = 0.4) +
  ylab("cvDNN") +
  xlab("")+
  ggtitle("cvSDNN for Baseline and Stressor")+
  theme(panel.background = element_blank())

#SDNN_plot


plot_grid(HR_plot, SDNN_plot + ylim(0, 30), labels = "AUTO")

```

## Respiration change baseline to stressor

``` {r echo = FALSE}

gg_resp <- data.frame(value  = c( resp$resp_baseline, resp$resp_stress),
                   category = c(rep('baseline', 29), rep('stress', 29)),
                   paired = c( 1:29, 1:29))


###plot of all ps for whom we have resp data - change in breathing rate from baseline to dungeon
resp_box <- ggplot(gg_resp, aes(x=category, y=value, fill=category)) +
  geom_boxplot(show.legend = F, coef = 0)+
  # geom_point() is used to make points at data values
  scale_fill_manual(values=cbPalette)+
  geom_point(show.legend = F)+
  # geom_line() joins the paired datapoints
  geom_line(aes(group=paired), alpha = 0.4) +
  ylab("Resp rate") +
  xlab("")+
  theme(panel.background = element_blank())


resp_box+ theme(axis.title=element_text(size=16)) + theme(axis.text =element_text(size=16)) 



```

# Respiratory analysis 

Correlation between the difference in respiration (stressor - baseline) and the difference in SDNN (stressor - baseline)


``` {r echo = FALSE}
################### RESPIRATORY ANALYSIS #############

###calculate adherence - deviation from 4 breaths per min 
adherence <- data.frame("p_id" = resp$p_id)
adherence$ad_training <- 4 - resp$resp_training
adherence$ad_stress <- 4 - resp$resp_stress
adherence$ad_training_stress_diff <- adherence$ad_stress - adherence$ad_training

##histogram of adherence scores for supplementary data 
#h <- ggplot(adherence, aes(x=ad_stress)) + 
#  geom_histogram(binwidth = 1.5) +
#  scale_x_continuous(breaks=seq(-14,2,by=2)) +
#  scale_y_continuous(breaks = seq(0, 14, by = 2)) + 
#  xlab("Resp adherence during stress" ) + 
#  theme_bw()


#h + theme(axis.title=element_text(size=16))

SDNN_resp <- SDNN[(SDNN$p_id %in% resp$p_id),] ##subset the SDNN dataframe by the ps who have resp
resp_SDNN <- resp[(resp$p_id %in% SDNN$p_id), ] ## make sure the breaths have corresponding cardio measures
HR_resp <- HR[(HR$p_id %in% resp$p_id),] ##ditto with the HR
adherence_resp <- adherence[(adherence$p_id %in% resp_SDNN$p_id),]

##merge together into one dataframe for the correlations/plots
resp_cardio <- merge(SDNN_resp, resp_SDNN, by = 'p_id')
resp_cardio <- merge(resp_cardio, HR_resp, by = 'p_id')
resp_cardio <- merge(resp_cardio, adherence_resp, by = "p_id")

rm(SDNN_resp, resp_SDNN, HR_resp, adherence_resp)

###corrlating the SDNN differences in stressor/baseline condition with resp differences in stress/bl
cor.test(resp_cardio$resp_stress_bl_diff, resp_cardio$SDNN_stress_BL_diff)
cor.test(resp_cardio$resp_stress_bl_diff, resp_cardio$HR_stress_BL_diff)

###plotting the correlation between resp rate and SDNN
ggscatter(resp_cardio, x = 'resp_stress_bl_diff', y = 'SDNN_stress_BL_diff',
          add = "reg.line", conf.int = TRUE,
          cor.coef = TRUE, cor.method = "pearson",
          xlab = "Difference in resp rate (dungeon - baseline)", ylab = "cvSDNN diff (dungeon - baseline)")

```

## Questionnaire analysis

``` {r echo = FALSE}

##sanity check - is there a difference in qualitative stressor/training stress ratings? 
t.test(quest$boat_score, quest$dungeon_score)

##make sure the rest of the questionnaires match up to the cardiac variables
quest <- quest[(quest$p_id %in% SDNN$p_id),]


####plotting the stress scores 
gg_stress <- data.frame(value  = c(quest$boat_score, quest$dungeon_score),
                        category = c(rep('training', 44), rep('stress', 44)),
                        paired = c( 1:44, 1:44))


ggplot(gg_stress, aes(x=reorder(category, value), y=value, fill=category)) +
  geom_boxplot(show.legend = F)+
  scale_fill_manual(values=cbPalette)+
  # geom_point() is used to make points at data values
  # geom_line() joins the paired datapoints
  ylab("Stress score (1-10)") +
  xlab("")+
  ggtitle("Stress scores for training and stress")+
  theme(panel.background = element_blank())
  
```
